AI Chatbots in K-12 Education: An Experimental Study of Socratic vs.~Non-Socratic Approaches and the Role of Step-by-Step Reasoning

### 2025-06-03 HSSC_Revised_Submission/main_report_anonym.tex
revised keep tex

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={AI Chatbots in K-12 Education: An Experimental Study of Socratic vs.~Non-Socratic Approaches and the Role of Step-by-Step Reasoning},
  pdfauthor={Andrea Blasco1,; Vicky Charisi1,2,},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{AI Chatbots in K-12 Education: An Experimental Study of Socratic vs.~Non-Socratic Approaches and the Role of Step-by-Step Reasoning}
\author{Andrea Blasco\textsuperscript{1,*} \and Vicky Charisi\textsuperscript{1,2,*}}
\date{This version: June 03, 2025}

@@

\setstretch{1.5}
\textsuperscript{1} European Commission, Joint Research Centre, Brussels, Belgium\\
\textsuperscript{2} Harvard University, Berkman Klein Center, Cambridge, MA, USA

\textsuperscript{*} Correspondence: \href{mailto:andrea.blasco@ec.europa.eu}{Andrea Blasco \textless{}\href{mailto:andrea.blasco@ec.europa.eu}{\nolinkurl{andrea.blasco@ec.europa.eu}}\textgreater{}}, \href{mailto:vcharisi@law.harvard.edu}{Vicky Charisi \textless{}\href{mailto:vcharisi@law.harvard.edu}{\nolinkurl{vcharisi@law.harvard.edu}}\textgreater{}}

@@

We recruited students after receiving ethical approval from the European Commission's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards. The data privacy protection protocol was approved by the European Commission's data protection officer. As an additional safeguard, given the minor age of the participants, we required their parents or legal guardians to provide us with written consent, allowing their children to participate in the study. The students also had to give their assent to participate by completing an online consent form.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/attitudes-AI-1.pdf}}
\caption{\label{fig:attitudes-AI}Student attitudes on AI in education}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/boot-accuracy-1.pdf}}
\caption{\label{fig:boot-accuracy}Comparison of Students' Absolute Error in Estimating Coin Jar Value with and without AI Explanations. All students received the AI-generated estimate of \$213 (the correct value was \$379.54). Still, those in the AI reasoning treatment also viewed the AI-generated step-by-step explanation for the estimation. Exposure to the AI-generated explanation significantly reduced the students' median absolute error.}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/perceived-accuracy-coins-1.pdf}}
\caption{\label{fig:perceived-accuracy-coins}Comparison of students' perceived accuracy of AI and human estimates across treatments on a five-point scale. Coefficients from separate ordinal logistic regressions controlling for students' location, gender, and prior experience with ChatGPT. Positive coefficients indicate increased perceived accuracy associated with students' exposure to AI reasoning.}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/messages-1.pdf}}
\caption{\label{fig:messages}Comparison of message frequency and word count per message across Socratic and non-Socratic treatments. Top panel (\textbf{A}) shows differences in the frequency of messages exchanged, while the bottom panels (\textbf{B} ) depict the word counts per message for the AI tutor (left) and the students (right).}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/confidence-1.pdf}}
\caption{\label{fig:confidence}This figure shows the impact of Socratic AI on students' confidence levels in their performance across three different tasks. Despite some differences between the Socratic and Non-Socratic groups, we found no significant association between the treatment assignment and students' declared confidence levels.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/confidence-regression-1.pdf}}
\caption{\label{fig:confidence-regression}This figure shows the coefficients from separate ordinal logistic regressions on students' confidence in their answer on a five-point scale, controlling for students' self-efficacy per task and task fixed effect. Results show no significant association between the treatment assignment and students' confidence levels.}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/helpfulness-1.pdf}}
\caption{\label{fig:helpfulness}This figure shows the impact of Socratic AI on students' perceived helpfulness of the AI tutor across three different tasks.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/helpful-regression-1.pdf}}
\caption{\label{fig:helpful-regression}The impact of Socratic AI on students' confidence levels in their performance across three different tasks. Despite some differences between the Socratic and Non-Socratic groups, we found no significant association between the treatment assignment and students' declared confidence levels}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/learning-1.pdf}}
\caption{\label{fig:learning}Effects on learning: \textbf{A} shows the \% of correct responses about the physics of sound propagation, illustrating the positive effect of interacting with the AI tutor, while no differences are associated with the Socratic AI. \textbf{B} shows the results of the verification question asking students to identify the fastest material for sound propagation (speed as meter per second reported) without AI assistance. Only 18\% responded accurately, indicating limited learning.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_files/figure-latex/interactions-1.pdf}}
\caption{\label{fig:interactions}Regression coefficients and 95\% confidence intervals from a linear regression using as dependent variable the (\textbf{A}) students' self-reported confidence level and (\textbf{B}) perceived helpfulness of the AI tutor. The controls include student's gender, grades, location, and ChatGPT experience. Treatment dummies are interacted with all the controls and the task type. All models also include individual student random effects, and the student's rated confidence in their skills before performing each task.}
\end{figure}

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={AI Chatbots in K-12 Education: An Experimental Study of Socratic vs.~Non-Socratic Approaches and the Role of Step-by-Step Reasoning},
  pdfauthor={Anonymous1,; Anonymous1,2,},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{AI Chatbots in K-12 Education: An Experimental Study of Socratic vs.~Non-Socratic Approaches and the Role of Step-by-Step Reasoning}
\author{Anonymous\textsuperscript{1,*} \and Anonymous\textsuperscript{1,2,*}}
\date{This version: June 03, 2025}

@@

\setstretch{1.5}
\textsuperscript{1} Anonymous Institute\\
\textsuperscript{2} Anonymous Institute

\textsuperscript{*} Correspondence: \href{mailto:Email}{Anonymous \textless{}Email\textgreater{}}, \href{mailto:Email}{Anonymous \textless{}Email\textgreater{}}

@@

We recruited students after receiving ethical approval from the Anonymous Institution's internal Ethical Review Board, ensuring that the consent procedures, data protection requirements, and experimental protocol complied with local laws and ethical research standards. The data privacy protection protocol was approved by the Anonymous Institution's data protection officer. As an additional safeguard, given the minor age of the participants, we required their parents or legal guardians to provide us with written consent, allowing their children to participate in the study. The students also had to give their assent to participate by completing an online consent form.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/attitudes-AI-1.pdf}}
\caption{\label{fig:attitudes-AI}Student attitudes on AI in education}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/boot-accuracy-1.pdf}}
\caption{\label{fig:boot-accuracy}Comparison of Students' Absolute Error in Estimating Coin Jar Value with and without AI Explanations. All students received the AI-generated estimate of \$213 (the correct value was \$379.54). Still, those in the AI reasoning treatment also viewed the AI-generated step-by-step explanation for the estimation. Exposure to the AI-generated explanation significantly reduced the students' median absolute error.}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/perceived-accuracy-coins-1.pdf}}
\caption{\label{fig:perceived-accuracy-coins}Comparison of students' perceived accuracy of AI and human estimates across treatments on a five-point scale. Coefficients from separate ordinal logistic regressions controlling for students' location, gender, and prior experience with ChatGPT. Positive coefficients indicate increased perceived accuracy associated with students' exposure to AI reasoning.}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/messages-1.pdf}}
\caption{\label{fig:messages}Comparison of message frequency and word count per message across Socratic and non-Socratic treatments. Top panel (\textbf{A}) shows differences in the frequency of messages exchanged, while the bottom panels (\textbf{B} ) depict the word counts per message for the AI tutor (left) and the students (right).}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/confidence-1.pdf}}
\caption{\label{fig:confidence}This figure shows the impact of Socratic AI on students' confidence levels in their performance across three different tasks. Despite some differences between the Socratic and Non-Socratic groups, we found no significant association between the treatment assignment and students' declared confidence levels.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/confidence-regression-1.pdf}}
\caption{\label{fig:confidence-regression}This figure shows the coefficients from separate ordinal logistic regressions on students' confidence in their answer on a five-point scale, controlling for students' self-efficacy per task and task fixed effect. Results show no significant association between the treatment assignment and students' confidence levels.}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/helpfulness-1.pdf}}
\caption{\label{fig:helpfulness}This figure shows the impact of Socratic AI on students' perceived helpfulness of the AI tutor across three different tasks.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/helpful-regression-1.pdf}}
\caption{\label{fig:helpful-regression}The impact of Socratic AI on students' confidence levels in their performance across three different tasks. Despite some differences between the Socratic and Non-Socratic groups, we found no significant association between the treatment assignment and students' declared confidence levels}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/learning-1.pdf}}
\caption{\label{fig:learning}Effects on learning: \textbf{A} shows the \% of correct responses about the physics of sound propagation, illustrating the positive effect of interacting with the AI tutor, while no differences are associated with the Socratic AI. \textbf{B} shows the results of the verification question asking students to identify the fastest material for sound propagation (speed as meter per second reported) without AI assistance. Only 18\% responded accurately, indicating limited learning.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{HSSC_Revised_Submission/main_report_anonym_files/figure-latex/interactions-1.pdf}}
\caption{\label{fig:interactions}Regression coefficients and 95\% confidence intervals from a linear regression using as dependent variable the (\textbf{A}) students' self-reported confidence level and (\textbf{B}) perceived helpfulness of the AI tutor. The controls include student's gender, grades, location, and ChatGPT experience. Treatment dummies are interacted with all the controls and the task type. All models also include individual student random effects, and the student's rated confidence in their skills before performing each task.}
\end{figure}
